Using functor categories to generate intermediate code with Agda

### 2025-04-20 draft.tex
definition almost completed; agda introduction added; files restructured

\documentclass[12pt,twoside,a4paper]{report}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{enumerate}
\usepackage[vmargin=2.5cm, hmargin=2.5cm]{geometry}
\usepackage{hyperref}
\usepackage{fontspec}
\usepackage{unicode-math}
\usepackage{stackengine}
\usepackage{minitoc}
\usepackage{amsthm}

@@

\graphicspath{{./fig/}}

@@

        Before other definitions, we need to get familiar with the notation of commutative diagrams.
        \subsubsection{Commutative diagrams}
        A \emph{diagram} in a category $\mathcal{C}$ is a directed graph whose vertices are $\mathcal{C}$-objects and whose edges are $\mathcal{C}$-morphisms.

@@

        \begin{figure}[htb!]
            \centering
            \begin{minipage}{0.3\textwidth}
                \centering
                \includegraphics[width=0.8\textwidth]{2_2_1_unit_laws.pdf}
                \caption{Commutative diagram for Unit Laws}
                \label{fig: unit_laws}
            \end{minipage} \hspace{0.2\textwidth}
            \begin{minipage}{0.3\textwidth}
                \centering
                \includegraphics[width=0.8\textwidth]{2_2_1_associativity_law.pdf}
                \caption{Commutative diagram for Associativity Law}
                \label{fig: associativity_law}
            \end{minipage}
        \end{figure}

@@

        \subsection{Cartesian closed category}
        \subsection{Functor}
        \begin{definition*}[Functor]
            Given two categories $\mathcal{C}$ and $\mathcal{D}$, a \emph{functor} $F: \mathcal{C} \to \mathcal{D}$ is specified by:
            \begin{itemize}
                \item 
                    a function 
                    \[\begin{aligned}
                        \textbf{obj}(\mathcal{C}) &\to \textbf{obj}(\mathcal{D}) \\
                        X &\mapsto F(X)
                    \end{aligned}\]

@@

            Given two categories $\mathcal{C}$ and $\mathcal{D}$, and two functors $F,G: \mathcal{C} \to \mathcal{D}$, a \emph{natural transformation} $\theta : F \to G$ is a family of morphisms $\theta_X \in \mathcal{D}{(F(X),G(X))}$ for each $X \in \textbf{obj}(\mathcal{C})$ such that for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, the following diagram 
            \begin{equation}
                \vcenter{\hbox{\includegraphics[width=0.3\textwidth]{2_2_3_natural_transformation.pdf}}}
                \label{cd: natural_transformation}
            \end{equation}

@@

    \section{Agda}
        \subsection{Standard library}

@@

        \subsection{Curry-Howard Correspondence}

        \subsection{Equality}

        \subsection{Proof Example}

        \subsection{Interactive programming with holes}

>>>>

\documentclass[12pt,twoside,a4paper]{report}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{enumerate}
\usepackage[vmargin=2.5cm, hmargin=2.5cm]{geometry}
\usepackage{hyperref}
\usepackage{fontspec}
\usepackage{unicode-math}
\usepackage{stackengine}
\usepackage{minitoc}
\usepackage{amsthm}
\usepackage{listings}

@@

\graphicspath{{./fig/category_theory/}}

@@

            \subsubsection{Opposite category}
            The idea of opposite category is that if we have a category $\mathcal{C}$, we can reverse the direction of all morphisms in $\mathcal{C}$ to obtain a new category $\mathcal{C}^{\textbf{op}}$.

            \begin{definition*}[Opposite category]
                Given a category $\mathcal{C}$, its \emph{opposite category} $\mathcal{C}^{\textbf{op}}$ is specified by:
                \begin{itemize}
                    \item 
                        the objects of $\mathcal{C}^{\textbf{op}}$ are the same as those of $\mathcal{C}$;
                    \item 
                        for each $X, Y \in \textbf{obj}(\mathcal{C})$, the morphisms from $X$ to $Y$ in $\mathcal{C}^{\textbf{op}}$ are the morphisms from $Y$ to $X$ in $\mathcal{C}$;
                    \item 
                        the identity morphism on $X$ in $\mathcal{C}^{\textbf{op}}$ is the identity morphism on $X$ in $\mathcal{C}$;
                    \item 
                        the composition of morphisms in $\mathcal{C}^{\textbf{op}}$ is defined as the composition of morphisms in $\mathcal{C}$.
                \end{itemize}
            \end{definition*}

                % \paragraph{The principle of duality}
                % Whenever one defines a concept or proves a theorem in category theory $\mathcal{C}$, one can obtain another concept or theorem in the opposite category $\mathcal{C}^{\textbf{op}}$, called the \emph{dual} of the original concept or theorem, by reversing all arrows in the original concept or theorem.

            The notation of commutative diagram is widely used in category theory as a convenient visual representation of the relationships between objects and morphisms in a category.
            \subsubsection{Commutative diagrams}
            A \emph{diagram} in a category $\mathcal{C}$ is a directed graph whose vertices are $\mathcal{C}$-objects and whose edges are $\mathcal{C}$-morphisms.

@@

            \begin{figure}[htb!]
                \centering
                \begin{minipage}{0.3\textwidth}
                    \centering
                    \includegraphics[width=0.8\textwidth]{unit_laws.pdf}
                    \caption{Commutative diagram for Unit Laws}
                    \label{fig: unit_laws}
                \end{minipage} \hspace{0.2\textwidth}
                \begin{minipage}{0.3\textwidth}
                    \centering
                    \includegraphics[width=0.8\textwidth]{associativity_law.pdf}
                    \caption{Commutative diagram for Associativity Law}
                    \label{fig: associativity_law}
                \end{minipage}
            \end{figure}

@@

        \subsection{Isomorphism}
        \begin{definition*}[Isomorphism]
            Given a category $\mathcal{C}$, a $C$-morphism $f : X \to Y$ is called an \emph{isomorphism} if there exists a $C$-morphism $g : Y \to X$ such that the following diagram commutes:
            \begin{equation}
                \vcenter{\hbox{\includegraphics[width=0.3\textwidth]{isomorphism.pdf}}}
                \label{cd: isomorphism}
            \end{equation}
            In other words, $f$ is an isomorphism if there exists a morphism $g$ such that $g \circ f = \textbf{id}_X$ and $f \circ g = \textbf{id}_Y$.

            The morphism $g$ is uniquely determined by $f$ and is called the \emph{inverse} of $f$, denoted as $f^{-1}$.

            Given two objects $X$ and $Y$ in a category $\mathcal{C}$, if there exists an isomorphism from $X$ to $Y$, we say that $X$ and $Y$ are \emph{isomorphic} in $\mathcal{C}$ and write $X \cong Y$.

        \end{definition*}
        [Considering adding examples of isomorphism]

        \subsection{Terminal object}
        \begin{definition*}[Terminal object]
            Given a category $\mathcal{C}$, an object $T \in \textbf{obj}(\mathcal{C})$ is called a \emph{terminal object} if for all $X \in \textbf{obj}(\mathcal{C})$, there exists a unique \textbf{C}-morphism $f : X \to T$.
        \end{definition*}
        [Considering adding examples of terminal object]

        Terminal objects are unique up to isomorphism. In other words, we have
        \begin{itemize}
            \item 
                if $T$ and $T'$ are both terminal objects in $\mathcal{C}$, then there exists a unique isomorphism $f : T \to T'$.
            \item
                if $T$ is a terminal object in $\mathcal{C}$ and $T \cong T'$, then $T'$ is also a terminal object in $\mathcal{C}$.
        \end{itemize}

        % \subsection{Initial object}
        % \begin{definition*}[Initial object]
        %     Given a category $\mathcal{C}$, an object $I \in \textbf{obj}(\mathcal{C})$ is called an \emph{initial object} if for all $X \in \textbf{obj}(\mathcal{C})$, there exists a unique \textbf{C}-morphism $f : I \to X$.
        % \end{definition*}
        % Initial object is dual to terminal object. By the principle of duality, initial objects are unique up to isomorphism. In other words, we have
        % \begin{itemize}
        %     \item 
        %         if $I$ and $I'$ are both initial objects in $\mathcal{C}$, then there exists a unique isomorphism $f : I \to I'$.
        %     \item
        %         if $I$ is an initial object in $\mathcal{C}$ and $I \cong I'$, then $I'$ is also an initial object in $\mathcal{C}$.
        % \end{itemize}

        \subsection{Binary product}
        \begin{definition*}[Binary product]
            Given a category $\mathcal{C}$, the \emph{binary product} of two objects $X$ and $Y$ in $\mathcal{C}$ is specified by
            \begin{itemize}
                \item a $\mathcal{C}$-object $X \times Y$;
                \item two $\mathcal{C}$-morphisms $\pi_1 : X \times Y \to X$ and $\pi_2 : X \times Y \to Y$ called the \emph{projections} of $X \times Y$;
            \end{itemize}
            such that for all $Z \in \textbf{obj}(\mathcal{C})$ and morphisms $f : Z \to X$ and $g : Z \to Y$, there exists a unique morphism $u : Z \to X \times Y$ such that the following diagram commutes in $\mathcal{C}$:
            \begin{equation}
                \vcenter{\hbox{\includegraphics[width=0.4\textwidth]{binary_product.pdf}}}
                \label{cd: binary_product}
            \end{equation}
            The unique morphism $u$ is written as 
            \[ \langle x, y \rangle : Z \to X \times Y \]
            where $x = \pi_1 \circ u$ and $y = \pi_2 \circ u$.
        \end{definition*}
        It can be shown that the binary product is unique up to (unique) isomorphism.

        \subsection{Exponential}
        \begin{definition*}[Exponential]
            Given a category $\mathcal{C}$ with binary products, the \emph{exponential} of two objects $X$ and $Y$ in $\mathcal{C}$ is specified by
            \begin{itemize}
                \item a $\mathcal{C}$-object $X \Rightarrow Y$;
                \item a $\mathcal{C}$-morphism $\textbf{app} : (X \Rightarrow Y) \times X \to Y$ called the \emph{application} of $X \Rightarrow Y$;
            \end{itemize}
            such that for all $Z \in \textbf{obj}(\mathcal{C})$ and morphisms $f : Z \times X \to Y$, there exists a unique morphism $u : Z \to X \Rightarrow Y$ such that the following diagram commutes in $\mathcal{C}$:
            \begin{equation}
                \vcenter{\hbox{\includegraphics[width=0.4\textwidth]{exponential.pdf}}}
                \label{cd: exponential}
            \end{equation}
            We write $\textbf{cur} f$ for the unique morphism $u$ such that $f = \textbf{app} \circ (\textbf{cur} f \times \textbf{id}_X)$, 
            where $\textbf{cur} f$ is called the \emph{currying} of $f$.
        \end{definition*}
        It can be shown that the exponential is unique up to (unique) isomorphism.

        \subsection{Cartesian closed category}
        \begin{definition*}[Cartesian closed category]
            A category $\mathcal{C}$ is called a \emph{Cartesian closed category} (ccc) if it has a terminal object, binary products and exponentials of any two objects.
        \end{definition*}

        \subsection{Functor}
        \begin{definition*}[Functor]
            Given two categories $\mathcal{C}$ and $\mathcal{D}$, a \emph{functor} $F: \mathcal{C} \to \mathcal{D}$ is specified by:
            \begin{itemize}
                \item 
                    a function 
                    \[\begin{aligned}
                        \textbf{obj}(\mathcal{C}) &\to \textbf{obj}(\mathcal{D}) \\
                        X &\mapsto F(X)
                    \end{aligned}\]

@@

            Given two categories $\mathcal{C}$ and $\mathcal{D}$, and two functors $F,G: \mathcal{C} \to \mathcal{D}$, a \emph{natural transformation} $\theta : F \to G$ is a family of morphisms $\theta_X \in \mathcal{D}{(F(X),G(X))}$ for each $X \in \textbf{obj}(\mathcal{C})$ such that for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, the following diagram 
            \begin{equation}
                \vcenter{\hbox{\includegraphics[width=0.3\textwidth]{natural_transformation.pdf}}}
                \label{cd: natural_transformation}
            \end{equation}

@@

    \section{Agda}
        \subsection{Basic datatypes and pattern matching}
        We will go through a simple example in PLFA \cite{plfa} to illustrate the basic datatypes and pattern matching in Agda.

@@

        \subsection{Curry-Howard-Lambek correspondence}

        \subsection{Equality, congruence and substitution}

        % \subsection{Proof Example}

        \subsection{Standard library}

        \subsection{Interactive programming with holes}
        A feature of Agda is that it allows us to write programs with holes interatively. A hole is a placeholder for a term that we have not yet defined. By leaving holes in place of undefined terms, we can write programs that are incomplete but still type-check, and Agda's type checker will guide completion of the program: the context window displays inferred types of the holes, available variables and candidate terms with their types. Holes also supports case split and refinement, which means we can fill in a hole partially and split it into smaller holes.

        [Considering adding examples of holes and case split, see https://plfa.github.io/Naturals/]

        In my implementation, I used holes to write the terms in the compiler. The complex terms are incrementally filled and verified by interatively refining partial implementations, reducing post-hoc debugging and ensuring robustness.
